Modeling the impact of IT- and AI-related emissions in a university setting

### 2025-09-09 doc/paper.tex
Nit
Increase university-owned laptop/monitor lifetimes per new central IT data, and adjust recommendations/conclusions.

\documentclass[11pt]{article}
\usepackage[margin=0.9in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{libertine}
\renewcommand{\sfdefault}{fvs} % Use as sans-serif
\renewcommand{\ttdefault}{fvm} % Use as monospace
\renewcommand{\ttfamily}{\small\fontfamily{\ttdefault}\selectfont} % Slightly smaller monospace font
\newcommand{\eg}{{\em e.g.}}
\newcommand{\coe}{CO$_2$e}
\newcommand{\gcoe}{g \coe}
\newcommand{\tcoe}{t \coe}
\newcommand{\kgcoe}{k\gcoe}
\newcommand{\gcoekwh}{\gcoe\ / kWh}
%\usepackage{helvet} % Helvetica font for sans-serif
%\usepackage{courier} % Courier font for monospace
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{
    hidelinks
}
\usepackage[style=numeric]{biblatex}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage{titling}

@@

\addbibresource{ref.bib}
\title{Modeling the impact of IT- and AI-related emissions in a university setting}
\author{Massimiliano Poletto\thanks{Independent consultant, max.poletto@gmail.com}, Nils Handler\thanks{University of Zürich, nils.handler@uzh.ch. Contact author.}}

@@

\begin{center}
  \begin{tabular}{|l|r|r|}
    \hline
    \textbf{Device type} & \textbf{Count} & \textbf{Life (y)}\\ \hline
    Laptops & 10,440 & 4 \\
    Monitors & 10,400 & 6 \\
    Servers & 1,121 & 5 \\
    Printers & 333 & 6 \\
    Mobile phones & 2,439 & 3 \\
    A/V devices & 431 & 6 \\ \hline
  \end{tabular}
  \label{tab:oncampusdevices}
\end{center}

@@

@@

Our assumptions about device duty cycle can be summarized as follows: (1) end-user devices are used during the day on weekdays, (2) servers run around-the-clock but at 80\% of peak power load, (3) mobile phones draw no power (they are unplugged) except for a few hours in the evening to charge, (4) printers are idle except for a few minutes per day. Standby power for all devices is negligible ($\sim 1$ W), except for printers which often draw 50--100 W.

@@

Using only the types of devices reported in the UZH 2024 Sustainability Report~\cite{uzh:sustainability:report}, and restricting embodied emissions to the Ecoinvent~\cite{ecoinvent} model used by that report, our model estimates emissions of 1,460 \tcoe, within 1\% of the 1,454 estimated in the UZH report, even though the methodology is somewhat different. Our model exposes average device power ratings and duty cycle, grid emissions intensity, etc., directly, whereas the approach used in the sustainability report relies on gross emissions factors (\kgcoe\ / device for embodied emissions, \kgcoe\ / device / day for operational emissions) provided by Ecoinvent.

@@

Hypothetically moving all UZH server workloads from on-premise data centers to a commercial cloud provider eliminates about 170 \tcoe\ of cooling overhead, and replaces 415 \tcoe\ of total server emissions with 380 \tcoe\ of cloud emissions. See Figure~\ref{fig:cloud}.

@@

For example, if we (1) massively reduce the set of end-user devices in scope to just those owned and managed by the university (about 10,000 laptops and monitors, 2,400 mobile phones), (2) triple the LLM per-query energy use to 3 Wh, and (3) assume that each of the 38,000 people at UZH uses a GPT-4o-class LLM 40 times per day, then emissions from AI still remain around 10\% of total (Figure~\ref{fig:sens1}).

@@

Only if we combine the ``low embodied emissions'' scenario with the ``pessimal LLM'' scenario does generative AI become a major factor, contributing 18\% of total emissions.

@@

Second, management of physical IT assets (workstations, monitors, etc.) is probably an underappreciated and underutilized lever for emissions reduction. Total emissions for these devices are dominated by manufacturing and disposal, so extending their life by a certain amount leads to an almost reciprocal reduction in emissions. For example, if we start the model with default parameters and then increase laptop lifetime from 4 to 6 years and mobile phone lifetime from 3 to 4 years, total campus emissions decrease from 3.9 k\tcoe\ to 3.2 k\tcoe, an 18\% reduction!

Note that an effective, large-scale extension of IT device lifetime is probably not just a matter of IT policy but of communication and education. The university could reach out to students and staff to sensitize them to these issues, provide more support for device maintenance and repair, etc.

@@

  \item Under the assumption that a large majority (80\%) of UZH students and staff have a laptop and smartphone, the manufacture of personal devices represents about $\frac{2}{3}$ of the university IT carbon footprint. Operational emissions account for only 15--20\% of total emissions of end-user devices (laptops, monitors, phones, etc.).

  \item Embodied emissions of end-user devices account for almost half of total emissoins even we consider only university-owned devices (equivalent to about $\frac{1}{4}$ of the population).

>>>>

\documentclass[11pt]{article}
\usepackage[margin=0.9in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{libertine}
\renewcommand{\sfdefault}{fvs} % Use as sans-serif
\renewcommand{\ttdefault}{fvm} % Use as monospace
\renewcommand{\ttfamily}{\small\fontfamily{\ttdefault}\selectfont} % Slightly smaller monospace font
\newcommand{\eg}{{\em e.g.}}
\newcommand{\coe}{CO$_2$e}
\newcommand{\gcoe}{g~\coe}
\newcommand{\tcoe}{t~\coe}
\newcommand{\kgcoe}{k\gcoe}
\newcommand{\gcoekwh}{\gcoe\ / kWh}
%\usepackage{helvet} % Helvetica font for sans-serif
%\usepackage{courier} % Courier font for monospace
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{
    hidelinks
}
\usepackage[style=numeric]{biblatex}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage{titling}

@@

\addbibresource{ref.bib}
\title{Modeling the impact of IT- and AI-related emissions in a university setting}
\author{Massimiliano Poletto\thanks{Independent consultant, max.poletto@gmail.com}, Nils Handler\thanks{University of Zürich, nils.handler@uzh.ch}}

@@

\begin{center}
  \begin{tabular}{|l|r|r|}
    \hline
    \textbf{Device type} & \textbf{Count} & \textbf{Life (y)}\\ \hline
    Laptops & 10,440 & 8 \\
    Monitors & 10,400 & 8 \\
    Servers & 1,121 & 5 \\
    Printers & 333 & 8 \\
    Mobile phones & 2,439 & 3 \\
    A/V devices & 431 & 8 \\ \hline
  \end{tabular}
  \label{tab:oncampusdevices}
\end{center}

@@

The device lifetime for university-managed laptops and monitors comes from UZH's central IT department and is remarkably long. For other types of devices we assume a similarly conservative management of hardware resources. We also estimate that student-owned laptops have a shorter lifetime of 5 years, and therefore, for the purposes of the results later in this paper, we set the default average lifetime of laptops in our model to 6 years.

@@

Our assumptions about device duty cycle can be summarized as follows: (1) end-user devices are used during the day on weekdays (250 days per year), (2) servers run around-the-clock but at 80\% of peak power load, (3) mobile phones draw no power (they are unplugged) except for a few hours in the evening to charge, (4) printers are idle except for a few minutes per day. Standby power for all devices is negligible ($\sim 1$ W), except for printers which often draw 50--100 W.

@@

Using only the types of devices reported in the UZH 2024 Sustainability Report~\cite{uzh:sustainability:report}, and restricting embodied emissions and mean device lifetimes to the Ecoinvent~\cite{ecoinvent} model used by that report, our model estimates emissions of 1,460 \tcoe, within 1\% of the 1,454 estimated in the UZH report, even though the methodology is somewhat different. Our model exposes average device power ratings and duty cycle, grid emissions intensity, etc., directly, whereas the approach used in the sustainability report relies on gross emissions factors (\kgcoe\ / device for embodied emissions, \kgcoe\ / device / day for operational emissions) provided by Ecoinvent.

@@

Hypothetically moving all UZH server workloads from on-premise data centers to a commercial cloud provider eliminates about 170~\tcoe\ of cooling overhead, and replaces 415~\tcoe\ of total server emissions with 380~\tcoe\ of cloud emissions. See Figure~\ref{fig:cloud}.

@@

For example, if we (1) massively reduce the set of end-user devices in scope to just those owned and managed by the university (about 10,000 laptops and monitors, 2,400 mobile phones), (2) triple the LLM per-query energy use to 3 Wh, and (3) assume that each of the 38,000 people at UZH uses a GPT-4o-class LLM 40 times per day, then emissions from AI are still just 11\% of total (Figure~\ref{fig:sens1}).

@@

Only if we combine the ``low embodied emissions'' scenario with the ``pessimal LLM'' scenario does generative AI become a substantial factor, contributing 16\% of total emissions.

@@

Second, management of physical IT assets (workstations, monitors, etc.) is probably an underappreciated and underutilized lever for emissions reduction. Total emissions for these devices are dominated by manufacturing and disposal, so extending their life by a certain amount leads to an almost reciprocal reduction in emissions. For example, if we start the model with default parameters and an industry-standard device lifetime of 4 years for laptops and 3 for mobile phones, then extend those to 6 and 4 years, respectively, total campus emissions decrease from 3.9 k\tcoe\ to 3.2 k\tcoe, an 18\% reduction! To its credit, the University of Zürich central IT department already keeps devices in service for about 8 years on average, so UZH may not have much room for improvement in this regard.

Note that an effective, large-scale extension of IT device lifetime is probably not just a matter of IT policy but of communication and education. Universities could reach out to students and staff to sensitize them to these issues, provide more support for device maintenance and repair, etc.

@@

  \item Under the assumption that a large majority (80\%) of UZH students and staff have a laptop and smartphone, the manufacture of personal devices represents almost $\frac{2}{3}$ of the university IT carbon footprint. Operational emissions account for only 15--20\% of total emissions of end-user devices (laptops, monitors, phones, etc.).

  \item Embodied emissions of end-user devices account for almost half of total emissions even we consider only university-owned devices (equivalent to about $\frac{1}{4}$ of the population).
